Why does female meiosis allow the opportunity for exploitation by selfish sperm? BETTER TITLE?

### 2013-09-14 writeup_and_notes/male_dependent_drive.tex

In the standard model of female meiotic drive, the driving allele at a balletic site is transmitted to the egg with probability  $d > \frac{1}{2}$, regardless of sperm genotype. 
%The standard model of female meiotic drive is two allele model, where the allele $1$ is transmitted $d$ fraction of the time through female meiosis ($d > \frac{1}{2}$), regardless of the genotype of the sperm. 
\yb{ To depict a case of a self-promoting (or `green-bearded') meiotic driver,  we modify this standard model such that the female meiotic driver is only effective when fertilized by a sperm carrying that allele (see Figure \ref{Eggsperm_2_allele_cartoon}).}
We then identify the conditions under which this self propagating driver can spread, and if it drive the evolution of suppressors? 
\yb{We note that the timing of fertilization relative to female meiosis places another constrain ton $d$, 
	for example, if fertilization (and therefore, sperm dependent drive) takes place at MII (as in mammals),
	female drive requires an uneven number of crossovers between the centromere and the drive locus, 
	so $d$ is bounded to be $<0.75$?.}  
\yb{I THINK THIS ORDER IS A BIT BETTER, BUT STILL ``OFF'', perhaps this is a footnote, or in methods or something...}
%In mammals, fertilization takes place at MII, so we imagine this drive must be taking place at MII in order
%for a sperm to have any influence. For drive to take place at MII there has to be an uneven number of
% crossovers between the centromere and the drive locus, such that realistically $d$ is bounded to be $<0.75$?. 
%In other systems where fertilization occurs during MI, sperm could influence either drive at MI or MII, 
%and drivers at MI can have a $d=1$ if they occur in tight linkage with the centromere.

@@

\yb{We first consider the situation where the driving allele is neutral in heterozygotes, 
	but deleterious in homozygotes (whose relative fitness is $1-s$). } 
%We first consider the situation where the driving allele is has no
%fitness consequences in heterozygotes, only in homozygotes (whose
%relative fitness is $1-s$). 
\yb{In this model, a standard meiotic driver can always invade because 
	when rare it occurs predominantly in heterozygotes and therefore it initially pays no fitness cost.} 
%A standard meiotic drive allele can always
%invade the population in this situation, as it is initially only
%present mostly in heterozygotes and rarely suffers the consequences of
%the problems in homozygotes. 
\yb{However, a severe homozygotes fitness cost ($s>XXd$) prevents  
	drivers from fixing  \citep{Prout1973}, 
	and provides an opportunity for the
	evolution of drive suppressors  \citep{XX}, corresponding well to 
	empirical examples of female meiotic drive. }

%However, if the cost in homozygotes is
%strong enough ($s>XXd$), then the allele is prevented from reaching
%fixation and instead is maintained as a balanced polymorphism in the
%population \citep{Prout1973}. Such balanced polymorphisms can then
%drive the evolution of supressors of drive \citep{XX}, and most of our
%examples of drive systems are of this form.

\yb{In contrast to a traditional driver, which can drive but pay no fitness cost when rare, 
	our green-bearded driver drives by creating homozygotes, and therefore pays a fitness cost even when rare. 
To simply increase when rare, our sperm-dependent driver must overcome a homozygous fitness cost.
Therefore with a recessive fitness cost,
	the conditions allowing the invasion of our green-bearded driver 
 	are far more restrictive than those for a standard meiotic driver (see Figure \ref{Invasion_space}).}

%Under a recessive fitness cost model our sperm dependent driver
%actually has far more restrictive conditions to invade the population,
%than a standard meiotic driver (see Figure \ref{Invasion_space}).
%That is because our allele which drives only when it is present in the
%fertilizing sperm is guarenteed to form a homzygote. Thus, 
%to even spread in the population when it is rare, our sperm-dependent
%driver has to overcome the cost it suffers in the homozygote state.

\yb{When a deleterious recessive green-bearded driver does spread, 
	it spends much time at low frequency, because when rare it is unlikely to meet a complimentary sperm. 
However, once common, a green bearded driver spreads rapidly to fixation due to their
	positive frequency dependent behavior (Inset in Figure \ref{Invasion_space}).  PERHAPS COMPARE TO TRADITIONAL DRIVER IN THIS FIGURE
Unlike standard meiotic drive systems, polymorphism of a green-bearded driver cannot be sustained by a homozygous cost, 
	because it already overcame this cost just to enter the population.
The lack of a balanced polymorphism precludes the evolution of an allele that suppresses this form of meiotic drive.}

%When conditions are suitable for such alleles to spread into the
%population, they spread slowly through low frequencies as they rarely
%drive as they are unlikely to be fertilized by a sperm with the right
%allele. Once they become established they spread rapidly to fixation due to their
%positive frequency dependent behavior (Inset in Figure \ref{Invasion_space}). 
%Unlike standard meiotic drive systems 
%these alleles can not become balanced in the population  by homozygote
%cost, as they have already over come this cost just to enter the population.

\yb{Inclusion of a heterozygous fitness cost (e.g. additivity) further constrains the evolution of a sperm-dependent driver. 
A  standard female  driver can invade from low frequency so long as drive in heterozygotes out weighs the fitness cost they suffer (XXX); 
	however, when at low frequency our sperm dependent driver pays the heterozygous fitness cost but rarely drives because it is rare in sperm.
Therefore, even very weak heterozygous fitness costs
	prevent the spread of a very rare green-bearded driver, unless its drive is very strong. 
With an additive cost, the dynamics of a green-bearded driver resembles that of an underdominant mutation (in that the system is bistable). 
Below a cutoff frequency the allele deterministically decreases in frequency 
	because it pays the  heterozygous fitness cost  but drives is ineffectively (see FigureS\ref{bistable_additive}). 
Above this frequency, the allele increases deterministically because the commonness of sperm carrying this allele 
	make drive effective enough to overcome the associated fitness cost.
%becayse heterozygote eggs are fertilized by sperm with our allele %sufficiently often to beat the costs. 
This bistability will  prevent sperm-dependent drivers from invading 	
	reasonably sized populations, and assures that if they do invade they will rapidly fix.}

%Things are even worse for the sperm-dependent driver if it suffers any
%fitness cost in the heterozygote. Our standard driver can invade the
%population from low frequency if its drive in heterozygotes out weighs
%the fitness cost they suffer (XXX). However, when at low frequency our
%sperm dependent driver is often in heterozygotes but will only drive
%if fertilized by the same allele. If our allele incurs even very weak costs
%in heterozygotes it can be prevented from spreading from very low
%frequencies, unless its drive is very strong. The dynamics of 
%our sperm-dependent drivers with an additive cost resemble an underdominant polymorphism
%(in that the system is bistable). Below a cutoff frequency the allele is
%unable to enter the population (in a deterministic model) as it is
%paying the cost of the drive but drive is ineffective (see Figure
%S\ref{bistable_additive}). Above this
%frequency the allele increases deterministically as its drive is sufficiently effective 
%becayse heterozygote eggs are fertilized by sperm with our allele
%sufficiently often to beat the costs. 
%This bistability suggests that sperm-dependent drivers will often be 
%unable to enter into reasonably sized populations, and if they do they
%should rapidly be spread fixation.

XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX
EDITED TO HERE
XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX

A model where the sperm allele affects outcome of female meiosis may not be biologically
realistic as there is very little sperm expression. More realistically the male genotype that produced the sperm
may be more relevant, as males could place products in their sperm that influenced the outcome of female meoisis.
However, in practice models that allow the influence of fertilizing male genotype on female meiosis 
seem to behave qualitatively to those based on the allele of the
fertilizing sperm (see Supplementary Figures XXX and XXX).

Given the difficulty that meotic drive alleles that promote themselves
through sperm have even entering the population, and the speed at
which they fix if they do, it seems very unlikely that such alleles
could drive the evolution of female

@@

>>>>

%yb version
In the standard model of female meiotic drive, the driving allele at a balletic site is transmitted to the egg with probability  $d > \frac{1}{2}$, regardless of sperm genotype. 
To depict a case of a self-promoting (or `green-bearded') meiotic driver,  we modify this standard model such that the female meiotic driver is only effective when fertilized by a sperm carrying that allele (see Figure \ref{Eggsperm_2_allele_cartoon}).
We then identify the conditions under which this self propagating driver can spread, and if it drive the evolution of suppressors? 
We note that the timing of fertilization relative to female meiosis places another constrain ton $d$, 
	for example, if fertilization (and therefore, sperm dependent drive) takes place at MII (as in mammals),
	female drive requires an uneven number of crossovers between the centromere and the drive locus, 
	so $d$ is bounded to be $<0.75$?.
\yb{I THINK THIS ORDER IS A BIT BETTER, BUT STILL ``OFF'', perhaps this is a footnote, or in methods or something...}

%coop version
%The standard model of female meiotic drive is two allele model, where the allele $1$ is transmitted $d$ fraction of the time through female meiosis ($d > \frac{1}{2}$), regardless of the genotype of the sperm. 
%In mammals, fertilization takes place at MII, so we imagine this drive must be taking place at MII in order
%for a sperm to have any influence. For drive to take place at MII there has to be an uneven number of
% crossovers between the centromere and the drive locus, such that realistically $d$ is bounded to be $<0.75$?. 
%In other systems where fertilization occurs during MI, sperm could influence either drive at MI or MII, 
%and drivers at MI can have a $d=1$ if they occur in tight linkage with the centromere.

@@

% YB VERSION
We first consider the situation where the driving allele is neutral in heterozygotes, 
	but deleterious in homozygotes (whose relative fitness is $1-s$).  
In this model, a standard meiotic driver can always invade because 
	when rare it occurs predominantly in heterozygotes and therefore it initially pays no fitness cost. 
However, a severe homozygotes fitness cost ($s>XXd$) prevents  
	drivers from fixing  \citep{Prout1973}, 
	and provides an opportunity for the
	evolution of drive suppressors  \citep{XX}, corresponding well to 
	empirical examples of female meiotic drive.

%COOP VERSION

%We first consider the situation where the driving allele is has no
%fitness consequences in heterozygotes, only in homozygotes (whose
%relative fitness is $1-s$). 
%A standard meiotic drive allele can always
%invade the population in this situation, as it is initially only
%present mostly in heterozygotes and rarely suffers the consequences of
%the problems in homozygotes. 
%However, if the cost in homozygotes is
%strong enough ($s>XXd$), then the allele is prevented from reaching
%fixation and instead is maintained as a balanced polymorphism in the
%population \citep{Prout1973}. Such balanced polymorphisms can then
%drive the evolution of supressors of drive \citep{XX}, and most of our
%examples of drive systems are of this form.

%YB VERSION
In contrast to a traditional driver, which can drive but pay no fitness cost when rare, 
	our green-bearded driver drives by creating homozygotes, and therefore pays a fitness cost even when rare. 
To simply increase when rare, our sperm-dependent driver must overcome a homozygous fitness cost.
Therefore with a recessive fitness cost,
	the conditions allowing the invasion of our green-bearded driver 
 	are far more restrictive than those for a standard meiotic driver (see Figure \ref{Invasion_space}).

%COOP VERSION
%Under a recessive fitness cost model our sperm dependent driver
%actually has far more restrictive conditions to invade the population,
%than a standard meiotic driver (see Figure \ref{Invasion_space}).
%That is because our allele which drives only when it is present in the
%fertilizing sperm is guarenteed to form a homzygote. Thus, 
%to even spread in the population when it is rare, our sperm-dependent
%driver has to overcome the cost it suffers in the homozygote state.

%YB VERSION
When a deleterious recessive green-bearded driver does spread, 
	it spends much time at low frequency, because when rare it is unlikely to meet a complimentary sperm. 
However, once common, a green bearded driver spreads rapidly to fixation due to their
	positive frequency dependent behavior (Inset in Figure \ref{Invasion_space}).  PERHAPS COMPARE TO TRADITIONAL DRIVER IN THIS FIGURE
Unlike standard meiotic drive systems, polymorphism of a green-bearded driver cannot be sustained by a homozygous cost, 
	because it already overcame this cost just to enter the population.
The lack of a balanced polymorphism precludes the evolution of an allele that suppresses this form of meiotic drive.

%COOP VERSION
%When conditions are suitable for such alleles to spread into the
%population, they spread slowly through low frequencies as they rarely
%drive as they are unlikely to be fertilized by a sperm with the right
%allele. Once they become established they spread rapidly to fixation due to their
%positive frequency dependent behavior (Inset in Figure \ref{Invasion_space}). 
%Unlike standard meiotic drive systems 
%these alleles can not become balanced in the population  by homozygote
%cost, as they have already over come this cost just to enter the population.

%YB VERSION
Inclusion of a heterozygous fitness cost (e.g. additivity) further constrains the evolution of a sperm-dependent driver. 
A  standard female  driver can invade from low frequency so long as drive in heterozygotes out weighs the fitness cost they suffer (XXX); 
	however, when at low frequency our sperm dependent driver pays the heterozygous fitness cost but rarely drives because it is rare in sperm.
Therefore, even very weak heterozygous fitness costs
	prevent the spread of a very rare green-bearded driver, unless its drive is very strong. 
With an additive cost, the dynamics of a green-bearded driver resembles that of an underdominant mutation (in that the system is bistable). 
Below a cutoff frequency the allele deterministically decreases in frequency 
	because it pays the  heterozygous fitness cost  but drives is ineffectively (see FigureS\ref{bistable_additive}). 
Above this frequency, the allele increases deterministically because the commonness of sperm carrying this allele 
	make drive effective enough to overcome the associated fitness cost.
%becayse heterozygote eggs are fertilized by sperm with our allele %sufficiently often to beat the costs. 
This bistability will  prevent sperm-dependent drivers from invading 	
	reasonably sized populations, and assures that if they do invade they will rapidly fix.

%COOP VERSION
%Things are even worse for the sperm-dependent driver if it suffers any
%fitness cost in the heterozygote. Our standard driver can invade the
%population from low frequency if its drive in heterozygotes out weighs
%the fitness cost they suffer (XXX). However, when at low frequency our
%sperm dependent driver is often in heterozygotes but will only drive
%if fertilized by the same allele. If our allele incurs even very weak costs
%in heterozygotes it can be prevented from spreading from very low
%frequencies, unless its drive is very strong. The dynamics of 
%our sperm-dependent drivers with an additive cost resemble an underdominant polymorphism
%(in that the system is bistable). Below a cutoff frequency the allele is
%unable to enter the population (in a deterministic model) as it is
%paying the cost of the drive but drive is ineffective (see Figure
%S\ref{bistable_additive}). Above this
%frequency the allele increases deterministically as its drive is sufficiently effective 
%becayse heterozygote eggs are fertilized by sperm with our allele
%sufficiently often to beat the costs. 
%This bistability suggests that sperm-dependent drivers will often be 
%unable to enter into reasonably sized populations, and if they do they
%should rapidly be spread fixation.

%yb version
%trying to make the models above sound less silly
While the sperm allele  could plausible influence the outcome of female meiosis, the limited levels of gene expression in sperm 
	suggest that sperm may influence the outcome of female meiosis as a function a the diploid genotype that produced the sperm perhaps due to proteins and RNAs packed in the sperm.
Models where male genotype, rather than sperm haplotype, determines the outcome of female meiosis 
	seem to behave qualitatively to those based on the allele of the
	fertilizing sperm (see Supplementary Figures XXX and XXX).
	\yb{THIS IS NOW WRONG WE NEED TO THINK ABOUT HOW TO WRITE THIS}

%coop version
%A model where the sperm allele affects outcome of female meiosis may not be biologically realistic as there is very little sperm expression. 
%More realistically the male genotype that produced the sperm
%may be more relevant, as males could place products in their sperm that influenced the outcome of female meoisis.
%However, in practice models that allow the influence of fertilizing male genotype on female meiosis 
%seem to behave qualitatively to those based on the allele of the
%fertilizing sperm (see Supplementary Figures XXX and XXX).

Given the difficulty that meiotic drive alleles that promote themselves
through sperm have entering the population, the speed at
which they fix if they do, \yb{and the absence of balanced polymorphisms at such loci,} it seems very unlikely that such alleles
could drive the evolution of female \yb{suppressors}.

@@

XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\newline
EDITED TO HERE\newline
XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\newline
